---
title: "Palmer Penguins Initial Analysis"
author: "Darvin Ojha"
format: html
editor: visual
execute:
  keep-md: true
---

## PalmerPenguinsAnalysis

This is an analysis of the Palmer's Penguin data set.

## Loading Packages and Dataset

Here we will load the tidyverse package and penguins data.

```{r}
#Load the tidyverse
library(tidyverse)
library(kableExtra)

#Read the penguins_samp1 data file from github
penguins <- read_csv("https://raw.githubusercontent.com/mcduryea/Intro-to-Bioinformatics/main/data/penguins_samp1.csv")

#See the first six rows of the data we've read in to our notebook
penguins %>% 
  head(2) %>% 
  kable() %>% 
  kable_styling(c("striped","hover"))
```

## About Our Data

The data we are working with is a dataset on Penguins, which includes 8 features measured on 44 penguins. The features included are physiological features (like bill length, bill depth, flipper length, body mass, etc) as well as other features like the year that the penguin was observed, the island the penguin was observed, the island the penguin was observed on, the sex of the penguin, and the species of the penguin.

## Interesting Questions to Ask

-   What is the average flipper length? What about for each species?

-   Are there more male or female penguins? What about per island or species?

-   What is the average body mass? What about by island? By species? By sex?

-   What is the ratio of bill length to bill depth for a penguin? What is the overall average of this metric? Does it change by species, sex, or island?

-   Does average body mass change by year?

    ## Data Manipulation Tools sand Strategies

We can look at individual columns in a data set or subsets of a column in a dataset. For example, if we are only interested in flipper length and species, we can `select()` those columns.

```{r}
penguins %>%
  select(species, body_mass_g)
```

If we want to `filter()` and only show certain rows, we can do that too.

```{r}
#We can filter by sex (catergorical variables)
penguins %>%
  filter(species =="Chinstrap")

#We can also filter by numerical varialbles
penguins %>%
  filter(body_mass_g >= 6000)

#We can filter 
penguins %>%
  filter((body_mass_g >=6000) | (island== "Torgersen"))
```

**Answering Our Questions**

Most of our questions involve summarizing data, and perhaps summarizing over groups. We can summarize data using the `summarize()` function, and group data using `group_by()`.

### Let's find the average flipper length.

```{r}
#Overall average flipper
penguins %>%
  summarize(avg_flipper_length = mean(flipper_length_mm))
#Single Species Average 
penguins %>%
  filter(species == "Gentoo") %>%
  summarize(avg_flipper_length = mean(flipper_length_mm))

#Grouped Average
penguins %>%
  group_by(species) %>%
  summarize(avg_flipper_length = mean(flipper_length_mm))
```

### How many of each species do we have?

```{r}
penguins %>%
  count(species)
```

### Let's find the number of male and female penguins.

How many of each sex are there? What about by island or species?

```{r}
#number of male penguins
penguins %>%
  count(sex)

penguins %>%
  group_by(species) %>%
  count(sex)


```

We can use a function called `mutate()` to add new columns to our data set.

```{r}
penguins_with_ratio <- penguins %>%
  mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm)

#Average Ratio
penguins %>%
   mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm) %>%
   summarize(mean_bill_ltd_ratio = mean(bill_ltd_ratio), 
          median_bill_ltd_ratio = median(bill_ltd_ratio))

#Average Ratio by Group
penguins %>%
  group_by(species) %>%
   mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm) %>%
   summarize(mean_bill_ltd_ratio = mean(bill_ltd_ratio), 
          median_bill_ltd_ratio = median(bill_ltd_ratio))


```

### Average body mass by year

```{r}
penguins %>%
  group_by(year) %>%
  summarize(mean_body_mass = mean(body_mass_g))
```

## Data Visualization

-   What is the distribution of penguin flipper lengths?

-   What is the distributive of penguin species?

-   Does the distribution of flipper length depend on the species of penguin?

    ```{r}
    penguins %>%
      ggplot() +
      geom_histogram( aes(x = flipper_length_mm),
                      bins = 15,
                      fill = "tan",
                      color = "Black") +
      labs(title = "Distribution of Flipper Length(mm)",
           y = "", x = "Flipper Length (mm)",
    subtitle = "Mean in Coral, Median in Blue") +
      geom_vline(aes(xintercept = mean
                 (flipper_length_mm)), lwd = 2, lty = "dashed", color = "coral"
      ) + 
                  geom_vline(aes(xintercept = median
                 (flipper_length_mm)), lwd = 2, lty = "dotted", 
                  color ="blue")

    ```

We will now look at the distribution of species.

```{r}
penguins%>%
  ggplot()+
  geom_bar(mapping = aes(x = species), color ="black", fill = "lightblue" )+
  labs(title = "Counts of Penguin Species",
      x="Species", y = "Count")
```

Let's make a scatter plot to see if bill length is correlated with bill depth.

```{r}
penguins%>%
  ggplot() +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm, color= species)) +
  geom_smooth(aes(x = bill_length_mm, y = bill_depth_mm, color= species), method = "lm")
```

A Final Question

We will find if the average bill length of a penguin exceeds 45mm or not.

```{r}
penguins%>%
  summarize(avg_bill_length = mean(bill_length_mm))
t.test(penguins$bill_length_mm, alternative= "greater" , mu= 45, conf.level= 0.95)
```
